---
title: "Exploratory Data Analysis"
date: "2024-02-07"
author: Nicholas Stevenson
output: html_document
---

  
This Quarto file loads the cleaned data and does some exploring.


As part of the exploratory analysis, you should produce plots or tables or other summary quantities for the most interesting/important quantities in your data. Depending on the total number of variables in your dataset, explore all or some of the others. Figures produced here might be histograms or density plots, correlation plots, etc. Tables might summarize your data.

Start by exploring one variable at a time. Then continue by creating plots or tables of the outcome(s) of interest and the predictor/exposure/input variables you are most interested in. If your dataset is small, you can do that for all variables. 

Plots produced here can be scatterplots, boxplots, violinplots, etc. Tables can be simple 2x2 tables or larger ones.

# Setup

```{r, message=FALSE, warning=FALSE, results='hide'}
#load needed packages. make sure they are installed.
library(here) #for data loading/saving
library(dplyr)
library(scales)
library(formattable)
library(skimr)
library(ggimage)
library(ggplot2)
library(readr)
library(readxl)
```


Load the data.

```{r}
#Path to data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","raw-data","NFL_data.xlsx")
#load data
nfl <- read_excel(data_location)
nfl
```


# Data exploration through tables

Showing a bit of code to produce and save a summary table.


```{r}
summary_df = skimr::skim(nfl)
print(summary_df)
# save to file
summarytable_file = here("results", "tables", "summarytable.rds")
saveRDS(summary_df, file = summarytable_file)
```

We are saving the results to the `results` folder. Depending on how many tables/figures you have, it might make sense to have separate folders for each. And/or you could have separate folders for exploratory tables/figures and for final tables/figures. Just choose a setup that makes sense for your project and works for you, and provide enough documentation that someone can understand what you are doing.


# Data exploration through figures

Bar plots for the continuous outcomes. 

Wins first.

```{r, message=FALSE}
p1 <- nfl %>% 
  ggplot(aes(x=Wins)) +
  geom_bar(fill="steel blue") 
plot(p1)
figure_file = here("results", "figures","Wins-distribution.png")
ggsave(filename = figure_file, plot=p1) 
```

Now Active Salary

```{r, message=FALSE}
p2 <- nfl %>% 
  ggplot(aes(x=reorder(Team, `ACTIVE53MAN`), y = `ACTIVE53MAN`)) +
  geom_bar(stat = "identity", fill="darkgreen") +
  coord_flip() +
  labs(title = "Active Salary by Team", x = "Team", y = "Active Salary")
plot(p2)

figure_file = here("results", "figures","Salary-distribution.png")
ggsave(filename = figure_file, plot=p2) 
```

# create median variables
```{r}
median_points_for <- median(nfl$Points_For)
median_points_for

median_points_against <- median(nfl$Points_Against)
median_points_against

median_wins <- median(nfl$Wins)
median_wins
```

# create images
```{r}
nfl$image_file <- paste0("team_images/", nfl$Team, ".png")
summary(nfl$image_file)
```

```{r}
points_for_wins <- nfl %>% 
  ggplot(aes(Wins,Points_For))+
  geom_point()+
  xlim(2,16)+
  ylim(175,575)+
  geom_hline(yintercept=365)+
  geom_vline(xintercept=9) +
  geom_image(aes(image = image_file), size = 0.03, asp = 4.0)

points_for_wins
```

```{r}
points_against_wins <- nfl %>% 
  ggplot(aes(Wins,Points_Against))+
  geom_point()+
  xlim(2,16)+
  ylim(175,575)+
  geom_hline(yintercept = 369)+
  geom_vline(xintercept = 9)+
  geom_image(aes(image=image_file), size = 0.03, asp = 4.0)

points_against_wins
```

```{r}
points_for_against <- nfl %>% 
  ggplot(aes(Points_For, Points_Against))+
  geom_point()+
  xlim(175,575)+
  ylim(175,575)+
  geom_hline(yintercept=369)+
  geom_vline(xintercept=365)+
  geom_image(aes(image = image_file), size = 0.03, asp = 4.0)

points_for_against
```

Plot wins as a function of salary
```{r, message=FALSE}
p3 <- nfl %>%
  ggplot(aes(x=`Wins`, y=`ACTIVE53MAN`))+
  geom_point() +
  geom_smooth(method='lm')
plot(p3)
figure_file = here("results", "figures", "Wins-Salary.png")
ggsave(filename = figure_file, plot=p3) 
```



For your own explorations, tables and figures can be "quick and dirty". As long as you can see what's going on, there is no need to polish them. That's in contrast to figures you'll produce for your final products (paper, report, presentation, website, etc.). Those should look as nice, polished and easy to understand as possible.


